High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-10-07 manuscript/Manuscript.tex
Updates from Overleaf

\begin{abstract}
New high-quality absolute paleointensity data from \textit{ca.} 1092 Ma anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$. This cooling rate corrected paleointensity estimate from Midcontinent Rift intrusions is similar to previous results from extrusive units that are dated to be between \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosite xenoliths can often have magnetic mineralogy dominated by grains with high coercivities, which can yield high success rates passing our selection criteria relative to their diabase hosts. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a strong geomagnetic field  persisted through 25 million years of the late Mesoproterozoic. These data are inconsistent with interpretations of continuous progressive decay of Earth's geodynamo strength through the Proterozoic Eon.

@@

\section*{Introduction}
%The question of the core and the importance of the observational data as complementary for numerical simulation work.

The strong geodynamo today is collectively powered by thermal convection and compositional convection in the liquid outer core. Studies have found that Earth may have sustained a geodynamo through the Archean \cite{Tarduno2014a}, and possibly as far back as the Hadean \cite{Tarduno2020a}. However, recent estimates of the thermal conductivity of the core suggest that the nucleation of the inner core nucleation could be $<$500 Ma, more recent than previous thought. Withouth the power of compositional convection provided by light elements being rejected into the outer core from the inner core, any identification of an earlier geodynamo would indicate that the core was stirred by other power sources, such as precipitation of Mg \cite{ORourcke2016a}, or perhaps
that the dynamo was generated by exotic processes like a convecting basal magma ocean \cite{Ziegler2013a}. In any case, a geodynamo solely driven by thermal conduction through the core-mantle boundary (CMB) prior to the nucleation of the inner core is likely to decay persistently due to heat loss.

That we have a solid inner core today is the result of secular cooling and differentiation of the layered Earth. The timing of inner core formation during Earth's thermal evolution is amongst the major unsolved questions in Earth's history. Due to the uncertain constraints on the thermal conductivity of the core \cite{Konopkova2016a, Gomi2013a, Ohta2016a}, numerical simulations have given drastically different predictions on the inner core nucleation age, ranging from about half a billion years ago to prior to the Archean \cite{Pozzo2012a, Gubbins2004a, Nimmo2015a}. Some models have suggested the thermal-convection-dominated liquid core motion experienced a long-term waning before the nucleation of the inner core, which kick-started the rigorous compositional convection between the solid inner core and liquid outer core \cite<e.g>{Davies2021a}. A corollary of such a prediction is that the geomagnetic axial dipole field, which is generated by the fluid motion in the liquid core, would have experienced a long-term decay before the emergence of the solid inner core, after which a significant increase in Earth's axial dipole field strength occurred \cite{Aubert2009a, Davies2021a}.

@@

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Fig. \ref{}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using the \textit{ca.} 1108 Ma Osler Volcanics, Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, and the Baraga and Marquette dikes, as well as the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Sprain2018a} obtained more high-quality paleointensity data using the \textit{ca.} 1108 Ma Osler Volcanics and the \textit{ca.} 1100 Ma Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years. In addition, an average VADM of 59 $\pm$ 11 ZAm$^2$ obtained \citeA{Kulakov2013a} using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics suggests that the present-day-like geomagnetic field strength likely persisted through the late stage of Midcontinent Rift magmatism.

Non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained from Midcontinent Rift rock units. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \siteA{Sprain2018a} who applied stricter paleointensity selection criteria and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. Despite the low success rate of paleointensity experiments, specimens that yield dominantly single-slope paleointensity results from the \textit{ca.} 1105 Ma flows of the Osler Volcanic Group, ca. 1095 Ma flows of the Mamainse Point volcanics \cite{Sprain2018a}, as well as the \textit{ca.} 1085 Ma flows of the Lake Shore Traps \cite{Kulakov2013a} suggest VADM values similar to that of the present-day geomagnetic field \cite{Sprain2018a}. Taken together, these rock units that have been dated to span $\sim$25 myr during the development of the Midcontinent Rift magmatism \cite{Swanson-Hysell2019a} may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

@@

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{PINT_BiCEP.pdf}
\centering
\caption{\footnotesize{Violin plot of paleointensity estimates using the bias corrected estimation of paleointensity (BiCEP) method \cite{Cych2021a}. }}
\label{fig:BiCEP}
\end{figure}

@@

% \begin{figure}
% \noindent\includegraphics[width=\textwidth]{pmag_plot_.pdf}
% \caption{\small{(A) Site-mean directions of anorthosite and diabase plotted on an equal-area plot. (B) Calculated VGPs from the anorthosite and diabase plotted in context of a previously synthesized \textit{ca.} 1.1 Ga Laurentia APWP from the Midcontinent Rift volcanics \citep{Swanson-Hysell2019a}. VGPs from both lithologies fall close to the 1095 Ma pole position. (C) Example Arai plots for diabase. Most diabase and many anorthosite specimens were rejected by selection criteria due to the zigzagging behavior shown in the diabase example plot. (D) Example Arai plots for anorthosite. A typical anorthosite specimen that passes selection shows straight Arai plot with a high paleointensity estimate before cooling rate correction. Note both specimens in (C) and (D) show dominantly single component magnetization in the inset orthogonal plots. The estimated field intensity is not cooling rate corrected.}}
% \label{fig:pmag}
% \end{figure}

>>>>

\begin{abstract}
New high-quality absolute paleointensity data from \textit{ca.} 1092 Ma anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$. This cooling rate corrected paleointensity estimate from Midcontinent Rift intrusions is similar to previous results from extrusive units that are dated to be between \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosite xenoliths can often have magnetic mineralogy dominated by grains with high coercivities, which can yield high success rates passing our selection criteria relative to their diabase hosts. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a strong geomagnetic field likely persisted through 25 million years of the late Mesoproterozoic. These data are inconsistent with interpretations of a progressive decay of Earth's geodynamo strength through the Proterozoic Eon.

@@

\section*{Introduction}

Earth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. This self-sustaining geodynamo is collectively powered by the heat flow across the core-mantle boundary (CMB) and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. Paleomagnetic studies have found that a dynamo may have existed by the Archean \cite{Tarduno2014a}, and possibly as far back as the Hadean \cite{Tarduno2020a}. However, some estimates of the thermal conductivity of the core suggest that the nucleation of the inner core could be younger than 1 Ga, more recent than previous thought \cite{Pozzo2012a, Gomi2013a, Labrosse2015a, Ohta2016a}. Without the power of compositional convection provided by the process of light elements being rejected into the outer core from the crystallizing inner core, the geodynamo was likely dominantly driven by thermal conduction through the CMB, with perhaps additional stirring by other power sources such Mg precipitation \cite{ORourcke2016a}, or perhaps there were intervals early in Earth history when the dynamo was generated by exotic processes such as a convecting basal magma ocean \cite{Ziegler2013a}. Long-term secular cooling of Earth has the potential to have decreased the power available to the geodynamo through time prior to the nucleation of the inner core \cite<e.g>{Davies2021a}. As a result, it is possible that strength of the geomagnetic axial dipole field would have experienced a long-term decay before the emergence of the solid inner core after which it would have strengthened \cite{Aubert2009a, Davies2021a}. However, the estimates for thermal conductivity values for the core continue to vary\cite{Konopkova2016a, Ohta2016a}. Obtaining an observational record of the timing of inner core formation would constrain the thermal evolution of the core and mantle, in addition to improving our understanding of the history of Earth’s dynamo.

@@

However, previous paleointensity data from intrusive and extrusive rock units in the Mesoproterozoic Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record Earth's virtual axial dipole moment (VADM) values similar to that of today, which are higher than the prediction of the long term decay curve (Fig. \ref{}). \citeA{Pesonen1983b} obtained an average virtual dipole moment of 93 ZAm$^2$ after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the Maimainse Point volcanics, intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flow of the \textit{ca.} 1096 Ma North Shore Volcanic Group. \citeA{Kulakov2013a} obtained an average VADM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by \citeA{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \citeA{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VADM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high VADM value estimate from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

@@

\subsection*{Bias Corrected Estimation of Paleointensity}
While using a set of selection criteria for excluding specimens with non-ideal paleointensity results is one common way of performing data analyses, the selection criteria themselves are often user-determined according to interpretations of certain groups of magnetic mineralogy on a case-by-case basis. The operation of performing a binary selection (accept or reject) on specimen results thus has an underlying assumption that the magnetic grain assemblage of specimens that pass the selection have a determined type of magnetic mineralogy. However, natural samples often have complex assemblage of magnetic carriers that span a range of composition, grain sizes and shapes. Thus, imposing a threshold value on certain parameters based on the paleointensity results can introduce bias from the analyzer. Alternatively, \citeA{Cych2021a} developed a Bias Corrected Estimation of Paleointensity (BiCEP) method that can eliminate biased introduced by using a set of arbitrarily defined selection criteria by taking all specimens into account for site-level paleointensity estimates. The method assumes that the paleointensity recorded by each specimen is biased away from a true value by an amount that is dependent on a single metric of nonlinearity (the curvature parameter $|\vec{k}|$; \citeA{Paterson2011a}) on the Arai plot. Although this method has the advantage of mitigating paleointensity estimate bias, it does not account for specimens that experience detrimental magnetic mineralogy alteration during heating steps unless such specimens are omitted before the analyses.

We apply the BiCEP method to all sites in this study and the cooling rate corrected site-level paleointensity estimates are shown in Figure \ref{fig:BiCEP}. For the anorthosite sites (e.g. AX11) that show good specimen-level consistency (Figure \ref{fig:PINT_cooling_corrected}) and straight specimen Arai plots (Figure \ref{fig:IZZI_examples}), the paleointensity estimates derived from the BiCEP method using all specimen data is indistinguishable with that derived from the site average value based on selected specimens (Figure \ref{fig:PINT_cooling_corrected}). For anorthosites that show rather large specimen-level dispersion (e.g. AX8), or diabase sites that have double-slop behaviors and significant MD component, the paleointensity estimates using the BiCEP method tend to have broader uncertainty ranges and the median values are lower than those from anorthosites that typically have more ideal paleointensity behaviors (Figure \ref{fig:PINT_cooling_corrected}). Overall, site mean value of the mean probability distribution of the BiCEP method results is consistent with the weighted mean value of the selection criteria-based estimates. In addition, the majority of the anorthosite sites give high probability estimate for there being a geomagnetic field value \textit{ca.} 1092 Ma at a paleolatitude of $\sim$30\textdegree similar to that of today. This result is consistent with the interpretation that the anorthosite xenoliths record a high geomagnetic field during the late Mesoproterozoic.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{code/code_output/AX_BD_BiCEP.pdf}
\centering
\caption{\footnotesize{Violin plot of paleointensity estimates from the anorthosite and diabase using the bias corrected estimation of paleointensity (BiCEP) method \cite{Cych2021a}. The results are grouped by individual cooling units following \citeA{Zhang2021b}. All estimates are corrected for cooling rate by a factor of 0.74. }}
\label{fig:BiCEP}
\end{figure}

@@
